% ── CHAPTER 9: CONVERGENCE 3: OBSERVER-DEPENDENCE (~6,000 words) ──
\chapter[Observer-Dependence]{Observer-Dependence: What You Can See Depends on What You Are}
\label{ch:observer-dependence}

% \epigraph{There is no view from nowhere.}{Thomas Nagel}

\firstword{T}{homas} Nagel's phrase has become something of a philosophical catchword, but it earns its keep.
There is a persistent temptation to imagine that reality, properly described, is described from no particular place --- that the truest account is the view an omniscient being would have, standing outside the universe, seeing everything at once.
In this imagined God's-eye view, every object has determinate properties; every event has a definite character; every fact is just a fact, independent of who is looking or what they are capable of registering.

The three traditions I am tracing in this book all reject that picture.
Not because they are anti-realist.
Not because they think the world is just whatever we say it is.
But because they have each, through their own methods and their own categories, arrived at the same structural conclusion: the way reality presents itself is inseparable from the kind of thing doing the looking.
The properties of a quantum system are irreducibly relational.
The depth of an actual occasion's experience is a function of its own relational complexity.
The kingdom of existence a creature inhabits determines what aspects of reality are available to it and what aspects remain entirely beyond its reach.

This is not relativism.
A thermometer's reading is genuinely real.
A person's reading is also genuinely real.
They are not the same reading.
Neither is false.
Neither is the complete truth.
Reality is one; the faces it presents to its various interlocutors depend on what those interlocutors are.

This chapter takes up that theme --- observer-dependence, hierarchy of comprehension, and what it means to say there is no view from nowhere --- across all three traditions.
The convergence here is subtler than the bare claim that observation affects what is observed.
It is the stronger claim that what becomes definite, what is even in principle available to a given type of observer, is a structural function of the observer's own relational constitution.
Not just that the measurement disturbs the system.
But that what is real for one kind of being is unavailable to another --- not because of ignorance, not because of imprecision, but because reality's architecture is hierarchical and perspectival all the way down.

\section{The Physics: No Absolute State, Only Relative States}
\label{sec:ch9-physics}

In 1996, the physicist Carlo Rovelli published a paper that has since become one of the most discussed and debated contributions to the foundations of quantum mechanics~\citep{Rovelli1996}.
The paper's title is ``Relational Quantum Mechanics,'' and its central claim is as plain as it is radical: quantum states are not absolute.
A quantum system does not have a definite quantum state, period.
It has a quantum state relative to some other system.
Change the relational context, and you change the state.

This sounds paradoxical until you remember that the same move is perfectly familiar in a different domain.
In special relativity, there is no such thing as an object's velocity, period.
There is only an object's velocity relative to some reference frame.
An electron at rest in my laboratory frame is moving at a substantial fraction of the speed of light in the frame of a muon created by a cosmic ray.
Neither description is wrong.
Neither is the ``real'' description.
Both are correct relative to their respective reference frames.
Velocity is frame-relative, not frame-independent.

Rovelli's proposal is that quantum states work the same way.
A particle's spin is not spin-up or spin-down.
It is spin-up relative to this detector, spin-down relative to that one.
A photon is not polarized along this axis or that one.
It is polarized along a given axis relative to a given polarizer.
There is no quantum state that the system simply has, independent of any relational context.

The radicalism of this move becomes clear when you push on it.
In ordinary quantum mechanics, the measurement problem asks: when does the wavefunction collapse?
Before measurement, the system is in a superposition of many possible states.
After measurement, it is in one definite state.
But when exactly does the collapse happen?
Many physicists have treated this as a deep puzzle about the nature of physical reality.

Relational quantum mechanics dissolves the puzzle by recasting it.
There is no measurement-independent state that collapses at some special moment.
There is only the state relative to a given observer.
When that observer interacts with the system, a definite outcome becomes actual for that observer.
But a second observer, who has not yet interacted with either the first observer or the system, may legitimately and correctly describe both the first observer and the system as still in superposition.

This sounds like a contradiction until you trace it carefully.
The two observers are not describing the same thing.
The first observer is describing the system's state relative to the first observer.
The second observer is describing the system-plus-first-observer's state relative to the second observer.
These are different relational contexts, and there is no inconsistency between them --- as long as the two observers have not yet interacted and compared notes.
Only when the second observer acquires information from the first does a definite outcome become actual for the second observer.

The key phrase is ``acquires information.''
In relational quantum mechanics, the event of one system acquiring information about another is itself a physical event.
It leaves a physical trace.
And that trace is what constitutes an outcome as definite for the acquiring system.
The measurement event does not disturb a pre-existing, determinate state.
It constitutes the outcome as definite, within that specific relational context.
The measurement is constitutive, not merely revelatory.

This is the feature I want to dwell on, because it will recur in the process-philosophical and \bahai accounts.
In a naive picture of measurement, the system has a definite state, and the measurement reads off what that state is.
In relational quantum mechanics, the system acquires a definite state only in the context of a particular relational interaction.
The outcome is not discovered; it is, in a precise sense, created by the relational event.
Created, and yet genuinely real.
Not invented.
Not arbitrary.
Constituted.

The philosopher-physicist Jenann Ismael has described this as the difference between a lock and a key~\citep{Rovelli1996}.
A pre-existing description would say: the key reveals the lock's shape.
A relational description says: the key and the lock co-determine a fitting or non-fitting through their interaction.
The fitting is real.
But it belongs to neither the key nor the lock in isolation.

Let me add one clarification that will matter for the convergence.
Relational quantum mechanics is not claiming that the world is subjective, or that different observers are trapped in incompatible and irreconcilable realities.
The constraint is consistency: if two observers have interacted (have entered into a relational context with each other), their descriptions of the shared outcomes must agree.
Rovelli is explicit that the relational facts are objective facts --- they are just facts about relations rather than facts about isolated substances.
The spin of an electron relative to a detector is a perfectly objective fact.
It is just not a fact about the electron alone; it is a fact about the electron-detector interaction.

Now consider what this implies about the limits of any observer's view.
A given observer can only directly access facts about systems that have interacted with that observer.
The totality of quantum fact is not, in principle, accessible to any single observer.
Not because some facts are hidden, but because facts are relational, and an observer's relational reach is finite.
There is always a wider relational context that exceeds what any particular observer can register.

That is the precise sense in which there is no view from nowhere.
Not that observers are systematically deceived.
Not that reality is unknowable.
But that every observer occupies a relational standpoint, and from every relational standpoint, there is more to reality than what that standpoint makes available.
The view from everywhere would require an entity in relational contact with everything.
No finite observer is that entity.
Reality is structured so that the whole always exceeds the reach of any particular perspective.

The analogy to special relativity is again instructive.
Special relativity does not say that velocity is unknowable or illusory.
It says velocity is frame-relative.
Every measurement of velocity is a perfectly objective measurement of a frame-relative quantity.
Relational quantum mechanics says the same about quantum properties.
Every measurement is a perfectly objective measurement of a relational quantity.
The view from nowhere would be a frame-independent velocity.
Physics has established that no such thing exists.

\section{The Process-Philosophical View: Grades of Experience}
\label{sec:ch9-process}

Whitehead arrived at a similar conclusion from a very different direction.
His starting point was not the quantum measurement problem but a puzzle about experience.
The dominant tradition in Western philosophy since Descartes had divided reality into two radically different kinds of substance: mind and matter.
Mind is characterized by subjectivity, by experience, by the felt quality of being something.
Matter is characterized by extension, by mechanism, by the absence of anything felt.

Whitehead found this division incoherent.
Not because he wanted to eliminate mind and reduce everything to mechanism --- that was the other, materialist horn of the same dilemma.
He found it incoherent because it required the emergence of experience from something entirely devoid of experience, at some unspecifiable threshold.
At what point does matter become aware?
At what level of neural complexity does mechanism become experience?
No one has given a satisfying answer, because the question presupposes a categorical gap that Whitehead thought was the source of the problem rather than a feature of the world.

His solution was to generalize experience downward rather than trying to explain how it emerges upward.
Every actual occasion, at every level of reality, has something like an experiential character --- a perspective, a grasping, a response to its environment.
This does not mean that electrons have rich conscious inner lives.
It means that the most rudimentary physical events exhibit the same formal structure that, at higher levels of integration, becomes what we recognize as experience: taking account of the environment, integrating incoming influences, and settling into a definite character through that process of relational reception.

Whitehead called this process prehension, from the Latin for grasping.
And he argued that different entities differ not in kind but in the depth and richness of their prehensive integration.
A simple physical occasion prehends its environment in a thin, predominantly physical, and largely unconscious way.
A complex living cell prehends its environment with more integration, more internal complexity, more capacity to respond selectively.
A vertebrate nervous system integrates prehensions at an enormously higher level of complexity, introducing something that begins to look like genuine experience.
A conscious human mind integrates at the highest level we know, producing the full richness of conscious thought, feeling, and creativity.

This is Whitehead's hierarchy of grades of experience, and it is structural, not merely taxonomic~\citep{Whitehead1933}.
It is not just a list of things in order of complexity.
It is a claim about the form of reality: that the capacity to take account of the world, to prehend and integrate the environment, is a fundamental feature of every actual entity --- and that this capacity varies, from the minimal in the most elementary physical occasions to the maximal in the most developed forms of consciousness.

Now here is the piece that connects to relational quantum mechanics.
Each actual occasion prehends its environment from its own spatiotemporal standpoint, with its own particular relational resources.
No occasion exhausts what it prehends.
No occasion takes in the full richness of the occasions it grasps.
There is always more to reality than any given prehension can integrate.

In \emph{Adventures of Ideas}, Whitehead is explicit about this~\citep{Whitehead1933}.
The finite perspective is not a defect of knowing; it is a structural feature of being.
To be an actual occasion is to be a finite relational perspective on reality.
The perspective is genuine.
The grasped reality is real.
But the perspective is partial, and its partiality is not an accident or a limitation to be eventually overcome.
It is what it means to be a particular entity rather than everything at once.

Higher grades of experience do not simply see more.
They integrate more, synthesizing a wider and richer relational field into a more complex and unified act of becoming.
A human mind does not simply have access to more facts than an amoeba.
It integrates its prehensions into a more deeply organized, more internally differentiated, more richly resonant unity.
The quantitative difference in reach becomes a qualitative difference in the character of experience.

Consider what this means for the claim that there is no view from nowhere.
In process ontology, there is no entity that prehends everything without being itself a particular entity with a particular perspective.
God, in Whitehead's framework, has a primordial nature that encompasses the full range of eternal objects --- the abstract forms of possibility --- but even God's consequent nature, the divine engagement with the actual world, takes in the world from God's own perspective and is enriched and shaped by the actual occasions that contribute to it.
The world is present to God in its fullness, but it is present as prehended, as taken in from a relational standpoint.

What Whitehead is insisting on is that the view from nowhere is not just practically unachievable.
It is formally incoherent.
To be an entity capable of knowing is to be an entity with a perspective.
The perspectiveless view would require an entity with no standpoint, no relational location, no particular position in the web of prehensive relations that constitutes reality.
Such an entity would not be a knower.
It would be nothing at all.

The hierarchy of grades is, from this angle, not a hierarchy of approximations to a view from nowhere.
It is a hierarchy of genuine perspectives, each valid at its own level, each disclosing aspects of reality that are inaccessible to the levels below it, and each failing to fully grasp what the levels above it can integrate.
The richest perspective is richer not because it approaches the view from nowhere, but because it integrates more of the world's relational complexity into a genuine, perspectival unity.

This maps onto relational quantum mechanics more cleanly than it might first appear.
Rovelli's different observers correspond, in broad structural terms, to Whitehead's actual occasions at different levels of relational complexity.
Each has a genuine perspective.
Each constitutes genuine facts through its relational interactions.
None exhausts the full reality it engages.
And the notion of a perspective-independent description of quantum reality is, in Rovelli's framework, as incoherent as the notion of a prehension-independent reality in Whitehead's.

\section{The Bahá'í View: The Hierarchy of Kingdoms}
\label{sec:ch9-bahai}

Among the most sustained treatments of observer-dependence in any spiritual tradition, the \bahai writings' account of the hierarchy of kingdoms is striking in its philosophical precision.
\AbdulBaha develops it at length in \emph{Some Answered Questions}, and the argument is not primarily about humility or the limits of human knowing.
It is an ontological argument about the structure of reality~\citep[Part 4]{SAQ}.

The passage deserves to be read carefully.
\AbdulBaha argues that each kingdom of existence --- mineral, vegetable, animal, human, and spiritual --- has its own mode of existence, its own form of participation in reality, its own capacity for perception and relation.
And he argues that each kingdom can, in principle, perceive and understand the kingdoms below it, but cannot perceive or understand the kingdom above it.

The mineral kingdom has its own form of being.
The laws of chemistry and crystallography that govern it are real laws, governing a real mode of existence.
But the mineral has no access to the mode of being that characterizes the vegetable kingdom.
It cannot perceive growth, nutrition, reproduction, or the sensitivity to environment that characterizes plant life.
Not because it lacks eyes or nervous systems --- those are higher-level instruments.
But because the mode of being proper to the vegetable kingdom is simply not available within the relational constitution of the mineral.

The vegetable kingdom, in turn, participates in a richer mode of being.
It does what the mineral does, but it also grows, reproduces, and responds to its environment in ways that the mineral cannot.
Yet the vegetable cannot perceive what is proper to the animal kingdom.
Voluntary movement, perception at a distance, the integration of multiple sensory channels into a unified representation of the environment --- these belong to a relational level that the vegetable's constitution does not reach.

The animal kingdom integrates still more.
It does what the vegetable does, and it adds the capacity for locomotion, perception, and something like memory and recognition.
But the animal cannot perceive what is proper to the human kingdom.
Abstract thought, language, mathematics, the capacity to consider possibilities that do not yet exist and to orient behavior toward a future that has not yet arrived --- these belong to a relational complexity that even the most sophisticated non-human animal does not reach.

And the human kingdom, for all its capacities, cannot perceive, in this life, what belongs to the spiritual realm.
The limitations here are not merely quantitative --- not merely that we perceive fewer things, less clearly.
They are structural.
The mode of being proper to the spiritual world is simply not available to an entity constituted as human beings in embodied life are constituted.

What is philosophically important here is the directionality of the blindness.
It is not symmetrical.
Higher kingdoms can understand lower kingdoms.
A biologist can fully describe the chemistry of a mineral.
A human being can understand the vegetable in terms of its chemistry and physiology.
A human being can, to a remarkable degree, understand the animal --- can model its behaviors, its nervous system, its cognition.
But neither the vegetable nor the animal can understand the human.

This asymmetry is not accidental.
It reflects the structure of the hierarchy itself.
Each level of existence subsumes the capacities of the levels below it while adding capacities that are irreducible to those lower levels.
A higher-level perspective genuinely integrates what lower-level perspectives can only partially encounter.
And the higher perspective is not simply a quantitative accumulation of lower-level information.
It is a qualitatively different mode of relational engagement with reality.

\AbdulBaha's point is not merely epistemic.
He is not saying only that higher kingdoms know more.
He is saying that they participate in a larger and richer relational reality.
The mineral is real.
The vegetable adds a dimension of reality that the mineral does not possess.
The animal adds another dimension.
The human adds another.
And the spiritual adds a dimension that exceeds, in principle, what human faculties in their current state can apprehend.

The \bahai writings also bring a crucial qualification that prevents this hierarchy from collapsing into mere relativism~\citep[para.~35]{SummonsLordHosts}.
The diversity of relational expressions does not mean that reality has no determinate character.
Reality is one.
Its face varies with the relational instrument through which it is perceived.
But the underlying reality is not fragmented or inconsistent.
``One single reality with manifold expressions owing to diversity of its instruments'' --- this is perspectival realism, not relativism.

The analogy the writings use elsewhere is illuminating: the same sun, many mirrors.
Each mirror reflects the sun truly.
None reflects it completely.
A small, rough mirror reflects a dim and distorted image.
A large, polished mirror reflects a clear and full image.
But both are reflecting the same sun, and no mirror, however polished, exhausts the sun itself.

The differences in the mirrors' reflections are real.
They are not errors.
They are accurate expressions of what the sun is, seen through these particular instruments.
And the fact that no mirror gives a complete reflection does not mean the sun is unknowable.
It means the sun exceeds any particular instrument's capacity.

This is not a claim about human subjectivity.
It is a claim about the architecture of reality.
The hierarchy of kingdoms is a hierarchy of relational instruments, each genuinely disclosing a real aspect of reality, none exhausting it.
The view from everywhere would require an instrument that subsumes all possible modes of being --- an instrument that is not itself at any particular level of the hierarchy, but that somehow includes them all.
That is precisely what the notion of absolute divine knowledge represents in the \bahai writings.
But for any finite mode of existence, reality presents a face proportioned to the instrument.

\section{The Structural Isomorphism: No View From Nowhere}
\label{sec:ch9-convergence}

It would be easy to note surface similarities between these three accounts and move on.
All three say that what you observe depends on how you observe, and stop there.
But that would miss the deeper structural claim, and it is the deeper structural claim that constitutes an isomorphism rather than a mere family resemblance.

All three traditions are making a common formal point.
Let me state it in abstract terms.

There is a set of relational entities.
Each entity has a particular relational constitution --- a particular position in the network of relations that constitutes reality.
From that relational constitution, the entity has access to certain facts about other entities in its environment.
Those facts are genuine facts, not subjective impressions.
They are constituted by the relational interaction between the entity and its environment.
But they are not all the facts.
There is always a wider relational context that exceeds what any particular relational constitution can access.

In relational quantum mechanics, the relational constitution is the observer's physical constitution --- the set of physical systems with which it has interacted and from which it can retrieve information.
In process philosophy, the relational constitution is the actual occasion's prehensive history --- the prior occasions it has integrated and the richness of its integrative capacity.
In the \bahai hierarchy, the relational constitution is the kingdom of existence --- the level of organizational complexity and relational integration that characterizes the entity's mode of being.

In each case, the facts that are accessible to a given relational constitution are genuine facts.
In each case, those facts are constituted by the relational interaction, not merely read off a pre-existing determinate state.
And in each case, there is always a wider context that exceeds the reach of any particular relational constitution.

The three traditions also converge on the implications of this structure.
None of them is relativist.
All three are perspectival realists.
They hold that perspectival access to reality is genuine access --- not a distortion or limitation but a real engagement with real features of the world.
The partiality of perspective is not a defect but a structural feature of being a finite relational entity in a relational world.

And all three are hierarchical in the same direction.
The hierarchy runs from simpler to richer relational constitution, from narrower to wider relational reach, from more restricted to less restricted access to the features of reality.
The hierarchy does not end at human consciousness.
In all three traditions, there is a sense in which the hierarchy extends beyond what current human capacities can fully grasp.

The common structure I am identifying is this: a relational ontology generates a hierarchy of perspectival access, and this hierarchy is not a defect to be corrected but a constitutive feature of what it means for reality to be relational.
The view from nowhere would be a perspective with no particular relational constitution --- an observer that is not any particular kind of observer.
All three traditions agree that no such entity exists, and that the very concept is incoherent.
What exists is a web of relational perspectives, each genuine, each partial, each constituting its own access to a reality that always exceeds any perspective's reach.

One further structural parallel deserves mention.
In relational quantum mechanics, the incompatibility between two observers' descriptions of the same system is not a contradiction --- it is expected and consistent, as long as the observers have not compared notes.
The incompatibility resolves, not by finding the ``true'' description, but by finding the description that is accurate for the joint system of both observers once they have interacted.
The hierarchical structure of \bahai kingdoms works similarly.
The mineral's description of reality and the human's description of reality are not contradictory.
They are each accurate for the level of relational constitution from which they emerge.
The apparent contradiction resolves not by declaring one wrong, but by recognizing that both are genuine expressions of a reality that exceeds either alone.

This is perspectival realism in its strongest form: the multiplicity of perspectives is not a sign of reality's incoherence but a sign of its inexhaustible depth.

\begin{convergencemoment}{Observer-Dependence: What You Can See Depends on What You Are}
In relational quantum mechanics~\citep{Rovelli1996}, a quantum system has no absolute, observer-independent state.
What becomes definite --- what is real in the sense of having a determinate value --- is constituted by the relational interaction between the system and an observer.
Two observers who have not compared notes may consistently assign different states to the same system.
Neither is wrong.
The observer-dependence is not a defect of quantum mechanics; it is a structural feature of relational reality.

In process philosophy~\citep{Whitehead1929,Whitehead1933}, every actual occasion prehends reality from its own relational standpoint, with its own particular prehensive resources.
No occasion exhausts the richness of what it prehends.
Whitehead's hierarchy of grades of experience is a hierarchy of relational perspectives, running from the most rudimentary physical occasions to the most complex forms of conscious integration.
Higher grades integrate more of the world's relational complexity into richer acts of becoming.
None integrates everything.

In the \bahai writings~\citep[Part 4]{SAQ}, the mineral cannot perceive what the vegetable perceives; the vegetable cannot perceive what the animal perceives; the animal cannot perceive what the human perceives; and the human cannot perceive, in this life, what belongs to the spiritual realm.
This hierarchy is not primarily about what different beings know; it is about what different beings are in relation to.
Each level of existence participates in a relational context that exceeds its comprehension.

One structure underlies all three.
What reality discloses depends on what kind of relational entity is doing the disclosing.
There is no view from nowhere --- not because reality is subjective, but because reality is relational, and every perspective on a relational world is itself a relational position.
The partiality of perspective is not a defect.
It is what it means to be a finite entity in an inexhaustible world.
\end{convergencemoment}
